---
title: "CCCC: ECMO and ICH patients"
author: "Adrian Barnett"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document: 
      reference_docx: rmarkdown-styles-SAP.docx
---
  
```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE, comment='', dpi=400)
options(width=1000) # Wide pages
options(scipen=999) # avoid scientific presentation
z = qnorm(0.975)
source('99_functions.R')
library(pander)
library(dplyr)
library(stringr)
library(tidyr)
library(flextable) # for nice tables
#library(broom)
library(summarytools)
# global options for summary tools
 st_options(plain.ascii = FALSE,       # This is very handy in all Rmd documents
            style = "rmarkdown",        # This too
            round.digits = 0, 
            headings = FALSE, 
            display.labels = FALSE,
            footnote = NA,             # Avoids footnotes which would clutter the results
            subtitle.emphasis = FALSE  # 
 ) 

# graphics things:
library(visdat) # for missing data
library(ggplot2)
library(ggupset)  # to plot combination of symptoms
g.theme = theme_bw() + theme(panel.grid.minor = element_blank())
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#999999", "#CC79A7")

## get the data
source = 'real_data'
#source = 'dummy_data'
if(source == 'dummy_data'){
  load(file='data/COVID_Data_stroke_dummy.RData') # from 0_read_data.R, dummy data
  # fixes with stroke
  patients = mutate(patients,
          stroke_group = ifelse(is.na(complication_stroke_date)==TRUE, 'None', stroke_group), # blank stroke group for those with no date
          stroke_group = ifelse(is.na(complication_stroke_date)==FALSE, sample(c('Hemorrhage','Ischemic','Other/Unknown','None'), size=n(), replace=TRUE), stroke_group) # randomly add stroke group for those with a date
                    )
}
if(source == 'real_data'){
  load(file='data/COVID_Data_stroke.RData') # from 0_read_data.R
}

# make last possible date based on database freeze
last_possible = as.Date(paste(data_date$day, data_date$month, '2020', sep='-'), format='%d-%b-%Y')

## data edits
# a) patients
patients = select(patients, 'pin', 'country','age', 'sex', 'ethnicity',
                  "height","weight",'date_admission','date_hospital_discharge',
                  'comorbidity_chronic_kidney_disease','comorbidity_severe_liver_disease',
                  'comorbidity_diabetes','comorbidity_hypertension',
                  'sofa','date_icu','date_ecmo','date_ecmo_discontinued',
                  'ecmo_type','ecmo_drainage_cannula_insertion_site',
                  'ecmo_return_cannula_insertion_site',
                  "ecmo_cardiac_arrest_before", "ecmo_worst_pa_o2_6hr_before",
                  "ecmo_worst_pa_co2_6hr_before", 'ecmo_continuous_rpt_before',
                  starts_with("complication_"), 'stroke_during_treatment','stroke_type', 'outcome') %>% # slim down variables
  # impossible admission date set to missing:
  mutate(date_admission = ifelse(date_admission >= last_possible, NA, date_admission), 
         date_admission = as.Date(date_admission, origin='1970-01-01'),
         # update stroke to add EOT
         complication_stroke = ifelse(!is.na(stroke_during_treatment) & stroke_during_treatment==1, 1, complication_stroke),
         # labels for factors
         complication_stroke = factor(complication_stroke, levels=0:1, labels=c('No','Yes')),
         complication_seizure = factor(complication_seizure, levels=0:1, labels=c('No','Yes')),
         complication_heart_failure = factor(complication_heart_failure, levels=0:1, labels=c('No','Yes')),
    comorbidity_chronic_kidney_disease = factor(comorbidity_chronic_kidney_disease, levels=0:1, labels=c('No','Yes')),
         comorbidity_severe_liver_disease = factor(comorbidity_severe_liver_disease, levels=0:1, labels=c('No','Yes')),
         comorbidity_diabetes = factor(comorbidity_diabetes, levels=0:1, labels=c('No','Yes')),
         comorbidity_hypertension = factor(comorbidity_hypertension, levels=0:1, labels=c('No','Yes')),
         ecmo_cardiac_arrest_before = factor(ecmo_cardiac_arrest_before, levels=0:1, labels=c('No','Yes')),
         ecmo_continuous_rpt_before = factor(ecmo_continuous_rpt_before, levels=0:1, labels=c('No','Yes')),
         any_ecmo = as.numeric(ecmo_type!=''),
         any_ecmo = factor(any_ecmo, levels=0:1, labels=c('No','Yes')))

# b) daily
daily = select(daily, pin, date_daily, ecmo, platelet_count, "aptt_aptr", "inr", alt_sgpt, "ast_sgot", 'bilirubin','blood_urea_nitrogen',
               "lymphocyte_count", 'eotd_anticoagulants', "eotd_anticoagulants_type",
               "pa_o2", "pa_co2",
               "eotd_haemorrhagic_complication", "eotd_haemorrhagic_complication_source",
               ) %>%
  mutate(date_daily = ifelse(date_daily >= last_possible, NA, date_daily), # impossible date set to missing
         date_daily = as.Date(date_daily, origin='1970-01-01')) 

## inclusion/exclusion ## 
patients = filter(patients,
    !country %in% c('Australia','United Kingdom')) # remove countries that did not collect stroke data
## exclusions
# age, must be adults
patients = filter(patients,
    !is.na(age),
    age >= 18) # 
# no ICU date
patients = filter(patients,
    !is.na(date_icu))
# missing sex (done below)

# combined
combined = full_join(patients, daily, by='pin')

# change 'missing' to NA
for_missing = select(patients, -pin, -any_ecmo, -stroke_type) %>% # remove ID; any_ecmo is never missing; stroke_type is a list variable
  mutate(
    sex = ifelse(sex=='Missing', NA,  sex),
    ethnicity = ifelse(ethnicity=='Missing', NA, ethnicity),
    outcome = ifelse(outcome %in% c('','unknown'), NA, outcome)) %>% # tidy up missing outcome
  mutate_if(~is.character(.x), ~ifelse(.x=='', NA, .x)) # change '' to missing for character variables

# reference dates for chloe
for_chloe = filter(patients, complication_stroke=="Yes") %>%
  select(pin, date_admission)
write.csv(for_chloe, file='data/for_chloe.csv', quote=FALSE, row.names = FALSE)

## exclusions
# ... none yet
```

This report uses the data from `r data_date$month` `r data_date$day`.

# How complete is the data?

#### a) Missing data from patient records

Here we look at patients missing stroke data to see how complete their other records are.

```{r, fig.width=8}
# remove conditional variables and lists
vis_dat(for_missing)
missing_gender = sum(is.na(for_missing$sex))
```

The grey areas are the missing data. 

#### b) Missing data from daily records

```{r, include=FALSE}
# exclude missing sex
to_exclude = filter(patients, sex=='Missing')$pin
combined = filter(combined, !pin %in% to_exclude)
patients = filter(patients, !pin %in% to_exclude)
daily = filter(daily, !pin %in% to_exclude)
for_missing = filter(for_missing, !is.na(sex))
```

```{r, fig.width=7}
# remove conditional variables and lists
for.missing = select(daily, -pin) %>%
  mutate_if(~is.character(.x), ~ifelse(.x=='', NA, .x)) # change '' to missing for character variables
vis_dat(for.missing)
```

The grey areas are the missing data.

#### Missing data and stroke

Do patients with a stroke have mortality data available? The table below shows whether the final outcome is known or unknown, and whether the stroke data is complete or missing. There is a lot of partially complete data. This uses the simple definition of stroke from the complications list. 

```{r, results='asis'}
to_table = mutate(patients, 
          outcome = ifelse(outcome=='', 'Outcome=Unknown', 'Outcome=Known'),
          stroke = ifelse(is.na(complication_stroke), 'Stroke=Missing', 'Stroke=Complete'))
tab_complete = with(to_table, ctable(outcome, stroke, chisq=FALSE, round.digits=0, dnn=c(' ',''), prop='t', totals=FALSE))
tab_complete
```


# Descriptive statistics

### Countries 

The table rows are ordered by frequency.

```{r, results='asis'}
tab = with(patients, freq(country, round.digits=0, report.nas = FALSE, order='freq'))
tab
```

### Numbers on ECMO

```{r, results='asis'}
tab_any = with(patients, freq(any_ecmo, round.digits=1, cumul = FALSE, report.nas	=FALSE))
tab_any
# percent and 95% CI
r = sum(patients$any_ecmo=='Yes')
n = nrow(patients)
p = r/n
se = sqrt(p*(1-p)/n)
lower = p - (z*se)
upper = p + (z*se)
```

The percent of patients experiencing any ECMO is `r roundz(p*100,1)`% with a 95% confidence interval from `r roundz(lower*100,1)`% to `r roundz(upper*100,1)`%.

### ECMO type

Just for those with any ECMO.

```{r, results='asis'}
to_table = filter(patients, ecmo_type!='')
tab_ecmo_type = with(to_table, freq(ecmo_type, round.digits=0, report.nas	=FALSE, cumul = FALSE))
tab_ecmo_type
```

### Cross-tabulation of sex by any ECMO

```{r, results='asis'}
tab_sex = with(patients, ctable(sex, any_ecmo, chisq=FALSE, round.digits=0, dnn=c(' ',''), prop='c'))
tab_sex
```

### Age summary statistics by any ECMO

```{r}
tab_age = group_by(patients, any_ecmo) %>%
  summarise(n=n(), 
            mean=round(mean(age, na.rm=TRUE)),
            sd=round(sd(age, na.rm=TRUE),1)
            )
f = flextable(tab_age)
autofit(f)
```


### Cross-tabulation of ethnicity by any ECMO

```{r, results='asis'}
tab_eth = with(patients, ctable(ethnicity, any_ecmo, round.digits=0, dnn=c(' ',''), prop='c'))
tab_eth
```

### Height and weight

##### Box & violin plots of height and weight

```{r, fig.width=8}
to_plot = select(patients, pin, any_ecmo, height, weight) %>%
  gather(key='var', value='res', -pin, -any_ecmo) %>%
  filter(!is.na(res))
vplot = ggplot(data=to_plot, aes(x=any_ecmo, y=res, fill=any_ecmo)) +
  geom_boxplot()+
  geom_violin(alpha=0.5)+
  facet_wrap(~var)+
  g.theme+
  xlab('')+
  ylab('Height or weight')+
#scale_y_continuous(breaks=NULL)+
  coord_flip()+
  scale_fill_manual('Any ECMO', values=cbPalette)
vplot
```

The black dots are potential outliers. The box contains the central 50% of the data.

Height is in cm and weight in kg.

##### Summary statistics for height and weight

```{r}
cont_vars = c('height','weight')
#
tab = select(patients, pin, all_of(cont_vars)) %>%
  tidyr::gather(key='Variable', value='Value', -pin) %>%
  group_by(Variable) %>%
  summarise(Missing = sum(is.na(Value)),
            Median = roundz(median(Value, na.rm=TRUE),0),
            Q1 = roundz(quantile(Value, probs=0.25, na.rm=TRUE),0),
            Q3 = roundz(quantile(Value, probs=0.75, na.rm=TRUE),0),
            IQR = paste(Q1 , ' to ', Q3, sep=''),
            Min = roundz(min(Value, na.rm=TRUE),0),
            Max = roundz(max(Value, na.rm=TRUE),0)) %>%
  select(Variable, Missing, Median, IQR, Min, Max) %>%
  mutate(Variable = ifelse(Variable=='age', 'Age', Variable))
ftab = flextable(tab) %>% 
  theme_box() %>%
  autofit()
ftab
```

### Cross-tabulation of chronic kidney disease by any ECMO

```{r, results='asis'}
tab_kidney = with(patients, ctable(comorbidity_chronic_kidney_disease, any_ecmo, chisq=FALSE, round.digits=0, dnn=c(' ',''), prop='c'))
tab_kidney
```

ECMO (No/Yes) is the columns.

### Cross-tabulation of severe liver disease by any ECMO

```{r, results='asis'}
tab_liver= with(patients, ctable(comorbidity_severe_liver_disease, any_ecmo, chisq=FALSE, round.digits=0, dnn=c(' ',''), prop='c'))
tab_liver
```

ECMO (No/Yes) is the columns.

### Cross-tabulation of diabetes by any ECMO

```{r, results='asis'}
tab_diab = with(patients, ctable(comorbidity_diabetes, any_ecmo, chisq=FALSE, prop='c', round.digits=0, dnn=c(' ','.')))
tab_diab
```

ECMO (No/Yes) is the columns.

### Cross-tabulation of hypertension by any ECMO

```{r, results='asis'}
tab_hyper = with(patients, ctable(comorbidity_hypertension, any_ecmo, chisq=FALSE, prop='c', round.digits=0, dnn=c(' ','.')))
tab_hyper
```
ECMO (No/Yes) is the columns.


### SOFA

##### Box & violin plots of SOFA

```{r}
vplot = ggplot(data=patients, aes(x=any_ecmo, y=sofa, fill=any_ecmo)) +
  geom_boxplot()+
  geom_violin(alpha=0.5)+
  g.theme+
  xlab(' ')+
  ylab('SOFA score')+
  scale_fill_manual('Any ECMO', values=cbPalette)
vplot
```

##### Summary statistics for SOFA

```{r}
cont_vars = c('sofa')
#
tab = select(patients, pin, all_of(cont_vars)) %>%
  tidyr::gather(key='Variable', value='Value', -pin) %>%
  group_by(Variable) %>%
  summarise(Missing = sum(is.na(Value)),
            Median = roundz(median(Value, na.rm=TRUE),0),
            Q1 = roundz(quantile(Value, probs=0.25, na.rm=TRUE),0),
            Q3 = roundz(quantile(Value, probs=0.75, na.rm=TRUE),0),
            IQR = paste(Q1 , ' to ', Q3, sep=''),
            Min = roundz(min(Value, na.rm=TRUE),0),
            Max = roundz(max(Value, na.rm=TRUE),0)) %>%
  select(Variable, Missing, Median, IQR, Min, Max) %>%
  mutate(Variable = ifelse(Variable=='age', 'Age', Variable))
ftab = flextable(tab) %>% 
  theme_box() %>%
  autofit()
ftab
```

## Combined table of percentages

The table is a combined table of the information above for the categorical variables shown as counts and percentages (in round brackets). 

```{r, results='asis'}
t1 = make_nice(tab_sex, rows = 'Female')
t2 = make_nice(tab_eth, rows = c('aboriginal','arab','black','east_asian','latin_american','south_asian','west_asian','white','other'))
t2[,1] = paste(t2[,1], '  ') # trying to play with alignment for gender
t3 = make_nice(tab_kidney, rows='Yes');  t3[1] = 'Kidney disease' # rename row
t4 = make_nice(tab_liver, rows='Yes');  t4[1] = 'Liver disease' 
t5 = make_nice(tab_diab, rows='Yes');  t5[1] = 'Diabetes' 
t6 = make_nice(tab_hyper, rows='Yes');  t6[1] = 'Hypertension' 
mega_table = data.frame(rbind(t1, t2, t3, t4, t5, t6))
row.names(mega_table) = NULL
names(mega_table) = c('Variable','Total','No ECMO','ECMO')
pander(mega_table, style='simple', justify='right')
```

# Summary statistics for patients on ECMO

```{r, include=FALSE}
ecmo_patients = filter(patients, any_ecmo=='Yes') %>%
  mutate(
    days_ecmo = as.numeric(date_ecmo_discontinued - date_ecmo + 0.5),
    days_hosp = as.numeric(date_hospital_discharge - date_admission + 0.5),
    day = as.numeric(date_ecmo - date_icu))
```


## Demographics

### Age

```{r}
cont_vars = c('age')
#
tab = select(ecmo_patients, pin, all_of(cont_vars)) %>%
  tidyr::gather(key='Variable', value='Value', -pin) %>%
  group_by(Variable) %>%
  summarise(Missing = sum(is.na(Value)),
            Median = roundz(median(Value, na.rm=TRUE),0),
            Q1 = roundz(quantile(Value, probs=0.25, na.rm=TRUE),0),
            Q3 = roundz(quantile(Value, probs=0.75, na.rm=TRUE),0),
            IQR = paste(Q1 , ' to ', Q3, sep=''),
            Min = roundz(min(Value, na.rm=TRUE),0),
            Max = roundz(max(Value, na.rm=TRUE),0)) %>%
  select(Variable, Missing, Median, IQR, Min, Max) %>%
  mutate(Variable = ifelse(Variable=='age', 'Age', Variable))
ftab = flextable(tab) %>% 
  theme_box() %>%
  autofit()
ftab
```

Age is only available in 5 year intervals.

### Categorical variables

```{r}
#
cat_vars = c('sex','ethnicity','country')
#
tab = select(ecmo_patients, pin, all_of(cat_vars)) %>%
  tidyr::gather(key='Variable', value='Value', -pin) %>%
  group_by(Variable, Value) %>%
  summarise(Count = n()) %>%
  group_by(Variable) %>%
  mutate(Percent = roundz(prop.table(Count)*100,0)) %>% # calculate percent
  arrange(Variable, -Count) %>% # arrange from high to low within variables
  ungroup() 
ftab = flextable(tab) %>%
  merge_v(j = c("Variable")) %>% # merge rows
  theme_box() %>%
  autofit()
ftab
```

The table rows are ordered by frequency.

### Summaries for time variables in days

```{r}
time_vars = c('days_ecmo','days_hosp','day')
#
tab = select(ecmo_patients, pin, all_of(time_vars)) %>%
  tidyr::gather(key='Variable', value='Value', -pin) %>%
  group_by(Variable) %>%
  mutate(Value = ifelse(Value < 0, NA, Value)) %>% # blank obvious errors
  summarise(Missing = sum(is.na(Value)),
            Median = roundz(median(Value, na.rm=TRUE),0),
            Q1 = roundz(quantile(Value, probs=0.25, na.rm=TRUE),0),
            Q3 = roundz(quantile(Value, probs=0.75, na.rm=TRUE),0),
            IQR = paste(Q1 , ' to ', Q3, sep=''),
            Min = roundz(min(Value, na.rm=TRUE),0),
            Max = roundz(max(Value, na.rm=TRUE),0)) %>%
  select(Variable, Missing, Median, IQR, Min, Max) %>%
  mutate(Variable = case_when(
    Variable == 'days_ecmo' ~ 'Length of the ECMO run',
    Variable == 'days_hosp' ~ 'Hospital length of stay',
    Variable == 'day' ~ 'Day put on ECMO'
  ))
ftab = flextable(tab) %>%
  theme_box() %>%
  autofit()
ftab
```


Just for the `r nrow(ecmo_patients)` on ECMO. IQR is the inter-quartile range.

## Other ECMO summaries 

```{r}
#
cat_vars = c('ecmo_drainage_cannula_insertion_site', 'ecmo_return_cannula_insertion_site', 'ecmo_cardiac_arrest_before', 'ecmo_continuous_rpt_before')
#
tab = select(ecmo_patients, pin, all_of(cat_vars)) %>%
  tidyr::gather(key='Variable', value='Value', -pin) %>%
  group_by(Variable, Value) %>%
  summarise(Count = n()) %>%
  group_by(Variable) %>%
  mutate(Percent = roundz(prop.table(Count)*100,0)) %>% # calculate percent
  arrange(Variable, -Count) %>% # arrange from high to low within variables
  ungroup() %>%
  mutate(
    # sort out missing
    Value = ifelse(is.na(Value)==TRUE, 'Missing', Value),
    Value = ifelse(Value=='', 'Missing', Value),
    # nicer labels
    Variable = ifelse(Variable=='ecmo_drainage_cannula_insertion_site', 'Drainage cannula insertion site', Variable),
         Variable = ifelse(Variable=='ecmo_return_cannula_insertion_site', 'Return cannula insertion site', Variable),
         Variable = ifelse(Variable=='ecmo_cardiac_arrest_before', 'Cardiac arrest before ECMO', Variable),
         Variable = ifelse(Variable=='ecmo_continuous_rpt_before', 'Continuous Renal Replacement Therapy before ECMO', Variable))
ftab = flextable(tab) %>%
  merge_v(j = c("Variable")) %>% # merge rows
  theme_box() %>%
  autofit()
ftab
```


Just for the `r nrow(ecmo_patients)` on ECMO. "(Empty string)" means the data are missing. 

### Blood gases before and after ECMO

#### a) PaO2

```{r, results='asis'}
pa_o2_data = blood_gases(before_var='ecmo_worst_pa_o2_6hr_before',
                       after_var = 'pa_o2') # from 99_functions.R
lplot = ggplot(data=pa_o2_data$to_plot, aes(x=xaxis, y=result, group=pin))+
  geom_line(col=grey(0.4))+
  geom_line(data=pa_o2_data$av_line, aes(x=xaxis, y=result, group=pin), col='dark red')+ # add average
  theme_bw()+
  scale_x_discrete(expand = c(0.05,0.05)) + # reduce white space
  xlab('')+
  ylab('PaO2 (log scale)')+
  scale_y_log10()
lplot
```

Here we examine the trends in PaO2 from 6 hours before ECMO initiation to the first PaO2 in the period after ECMO initiation.

The line plot shows the individual change in PaO2 for `r length(unique(pa_o2_data$to_plot$pin))` patients with available blood gas data and ECMO dates. The y-axis is on a log-scale because of the positive skew in PaO2. The red lines shows the average difference between the two times.

The mean percentage change between the worst before reading and first reading after ECMO was `r roundz(pa_o2_data$test$pdiff, 1)` with a 95% confidence interval from `r roundz(pa_o2_data$test$conf.int[1], 1)` to `r roundz(pa_o2_data$test$conf.int[2], 1)` and p-value `r format.pval(pa_o2_data$test$p.value, eps=0.001, digits=3)` (using a paired t-test with log-transformed data).

#### b) PaCO2

```{r, results='asis'}
pa_co2_data = blood_gases(before_var='ecmo_worst_pa_co2_6hr_before',
                       after_var = 'pa_co2') # from 99_functions.R
lplot = ggplot(pa_co2_data$to_plot, aes(x=xaxis, y=result, group=pin))+
  geom_line(col=grey(0.4))+
  geom_line(data=pa_co2_data$av_line, aes(x=xaxis, y=result, group=pin), col='dark red')+ # add average
  scale_x_discrete(expand = c(0.05,0.05)) + # reduce white space
  theme_bw()+
  xlab('')+
  ylab('PaCO2 (log-scale)')+
  scale_y_log10()
lplot
```

Here we examine the trends in PaO2 from 6 hours before ECMO initiation to the first PaCO2 in the period after ECMO initiation.

The line plot shows the individual change in PaCO2 for `r length(unique(pa_o2_data$to_plot$pin))` patients with available blood gas data and ECMO dates. The y-axis is on a log-scale because of the positive skew in PaCO2.

The mean percentage change between the worst before reading and first reading after ECMO was `r roundz(pa_co2_data$test$pdiff, 1)` with a 95% confidence interval from `r roundz(pa_co2_data$test$conf.int[1], 1)` to `r roundz(pa_co2_data$test$conf.int[2], 1)` and p-value `r format.pval(pa_co2_data$test$p.value, eps=0.001, digits=3)` (using a paired t-test with log-transformed data).



# Values collected during the ICU stay

```{r, include=FALSE}
# data to plot, must have dates
to_plot = filter(combined, 
                 !is.na(date_icu),
                 !is.na(date_daily)) %>%
  mutate(day = as.numeric(date_daily - date_icu)) %>%
  filter(day <= 21) # exclude long stays for the plot
```

I limit the plots below to the first 21 days during the ICU stay.

## Platelet count

##### Boxplot over time

```{r, fig.width=8}
tplot = ggplot(data=to_plot, aes(x=factor(day), y=platelet_count, col=factor(any_ecmo)))+
  geom_boxplot()+
  xlab('Day in ICU')+
  ylab('Platelet count')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}
to_stat = group_by(to_plot, day, any_ecmo) %>%
  summarise(mean = mean(platelet_count, na.rm = TRUE),
            se = sem(platelet_count)) %>%
  ungroup() %>%
  mutate(lower = mean - (z*se),
         upper = mean + (z*se),
         day = ifelse(any_ecmo=='Yes', day+0.2, day)) # jitter x slightly to avoid overlap
tplot = ggplot(data=to_stat, aes(x=day, y=mean, ymin=lower, ymax=upper, col=factor(any_ecmo)))+
  geom_point()+
  geom_errorbar(width=0)+
  xlab('Day in ICU')+
  ylab('Platelet count')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

Just for the first 21 days and the `r length(unique(to_plot$pin))` patients in ICU.

## Blood urea nitrogen

##### Boxplot over time

```{r, fig.width=8}
tplot = ggplot(data=to_plot, aes(x=factor(day), y=blood_urea_nitrogen, col=factor(any_ecmo)))+
  geom_boxplot()+
  xlab('Day in ICU')+
  ylab('Blood urea nitrogen')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```


##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}
to_stat = group_by(to_plot, day, any_ecmo) %>%
  summarise(mean = mean(blood_urea_nitrogen, na.rm = TRUE),
            se = sem(blood_urea_nitrogen)) %>%
  ungroup() %>%
  mutate(lower = mean - (z*se),
         upper = mean + (z*se),
         day = ifelse(any_ecmo=='Yes', day+0.2, day)) # jitter x slightly to avoid overlap
tplot = ggplot(data=to_stat, aes(x=day, y=mean, ymin=lower, ymax=upper, col=factor(any_ecmo)))+
  geom_point()+
  geom_errorbar(width=0)+
  xlab('Day in ICU')+
  ylab('Blood urea nitrogen')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

Just for the first 21 days and the `r length(unique(to_plot$pin))` patients in ICU.

## Bilirubin

##### Boxplot over time

```{r, fig.width=8}
tplot = ggplot(data=to_plot, aes(x=factor(day), y=bilirubin, col=factor(any_ecmo)))+
  geom_boxplot()+
  xlab('Day in ICU')+
  ylab('Bilirubin')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```


##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}
to_stat = group_by(to_plot, day, any_ecmo) %>%
  summarise(mean = mean(bilirubin, na.rm = TRUE),
            se = sem(bilirubin)) %>%
  ungroup() %>%
  mutate(lower = mean - (z*se),
         upper = mean + (z*se),
         day = ifelse(any_ecmo=='Yes', day+0.2, day)) # jitter x slightly to avoid overlap
tplot = ggplot(data=to_stat, aes(x=day, y=mean, ymin=lower, ymax=upper, col=factor(any_ecmo)))+
  geom_point()+
  geom_errorbar(width=0)+
  xlab('Day in ICU')+
  ylab('Bilirubin')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

Just for the first 21 days and the `r length(unique(to_plot$pin))` patients in ICU.

## APTT/APTR

##### Boxplot over time

```{r, fig.width=8}
tplot = ggplot(data=to_plot, aes(x=factor(day), y=aptt_aptr, col=factor(any_ecmo)))+
  geom_boxplot()+
  xlab('Day in ICU')+
  ylab('APTT/APTR')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}
to_stat = group_by(to_plot, day, any_ecmo) %>%
  summarise(mean = mean(aptt_aptr, na.rm = TRUE),
            se = sem(aptt_aptr)) %>%
  ungroup() %>%
  mutate(lower = mean - (z*se),
         upper = mean + (z*se),
         day = ifelse(any_ecmo=='Yes', day+0.2, day)) # jitter x slightly to avoid overlap
tplot = ggplot(data=to_stat, aes(x=day, y=mean, ymin=lower, ymax=upper, col=factor(any_ecmo)))+
  geom_point()+
  geom_errorbar(width=0)+
  xlab('Day in ICU')+
  ylab('APTT/APTR')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

Just for the first 21 days and the `r length(unique(to_plot$pin))` patients in ICU.

## INR

##### Boxplot over time

```{r, fig.width=8}
tplot = ggplot(data=to_plot, aes(x=factor(day), y=inr, col=factor(any_ecmo)))+
  geom_boxplot()+
  xlab('Day in ICU')+
  ylab('INR')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```


##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}
to_stat = group_by(to_plot, day, any_ecmo) %>%
  summarise(mean = mean(inr, na.rm = TRUE),
            se = sem(inr)) %>%
  ungroup() %>%
  mutate(lower = mean - (z*se),
         upper = mean + (z*se),
         day = ifelse(any_ecmo=='Yes', day+0.2, day)) # jitter x slightly to avoid overlap
tplot = ggplot(data=to_stat, aes(x=day, y=mean, ymin=lower, ymax=upper, col=factor(any_ecmo)))+
  geom_point()+
  geom_errorbar(width=0)+
  xlab('Day in ICU')+
  ylab('INR')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

Just for first 21 days and the `r length(unique(to_plot$pin))` patients in ICU.

## ALT-SGPT

For this variable I use a log-transform because the variable has a very strong positive skew.

##### Boxplot over time

```{r, fig.width=8}
tplot = ggplot(data=to_plot, aes(x=factor(day), y=alt_sgpt, col=factor(any_ecmo)))+
  geom_boxplot()+
  xlab('Day in ICU')+
  ylab('ALT-SGPT (log-scale)')+
  scale_y_log10()+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

The y-axis is on a log scale.

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}
to_stat = group_by(to_plot, day, any_ecmo) %>%
  summarise(mean = mean(log(alt_sgpt), na.rm = TRUE),
            se = sem(log(alt_sgpt))) %>%
  ungroup() %>%
  mutate(lower = mean - (z*se),
         upper = mean + (z*se),
         day = ifelse(any_ecmo=='Yes', day+0.2, day)) # jitter x slightly to avoid overlap
tplot = ggplot(data=to_stat, aes(x=day, y=mean, ymin=lower, ymax=upper, col=factor(any_ecmo)))+
  geom_point()+
  geom_errorbar(width=0)+
  xlab('Day in ICU')+
  ylab('Log-transformed ALT-SGPT')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

Just for the first 21 days and the `r length(unique(to_plot$pin))` patients in ICU.

The y-axis is for the log-transformed value.

## AST-SGOT

For this variable I use a log-transform because the variable has a very strong positive skew.

##### Boxplot over time

```{r, fig.width=8}
tplot = ggplot(data=to_plot, aes(x=factor(day), y=ast_sgot, col=factor(any_ecmo)))+
  geom_boxplot()+
  xlab('Day in ICU')+
  ylab('AST-SGOT (log-scale)')+
  scale_y_log10()+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

The y-axis is on a log scale.

##### Plot of means and 95% confidence intervals for the means over time

```{r, fig.width=8}
to_stat = group_by(to_plot, day, any_ecmo) %>%
  summarise(mean = mean(log(ast_sgot), na.rm = TRUE),
            se = sem(log(ast_sgot))) %>%
  ungroup() %>%
  mutate(lower = mean - (z*se),
         upper = mean + (z*se),
         day = ifelse(any_ecmo=='Yes', day+0.2, day)) # jitter x slightly to avoid overlap
tplot = ggplot(data=to_stat, aes(x=day, y=mean, ymin=lower, ymax=upper, col=factor(any_ecmo)))+
  geom_point()+
  geom_errorbar(width=0)+
  xlab('Day in ICU')+
  ylab('Log-transformed AST-SGOT')+
  scale_color_manual('Any ECMO', values=c('darkorange1','cyan4'))+
  g.theme
tplot
```

Just for the first 21 days and the `r length(unique(to_plot$pin))` patients in ICU.

The y-axis is for the log-transformed value.

## Anticoagulants in the last 24 hours

```{r, fig.width=8}
to_plot = mutate(to_plot, facet = paste("Any ECMO = ", any_ecmo, sep=''), # nice facet
                 eotd_anticoagulants_plot = ifelse(is.na(eotd_anticoagulants), 2, eotd_anticoagulants)) # replace missing %>%
bplot = ggplot(data=to_plot, aes(x=factor(day), fill=factor(eotd_anticoagulants_plot)))+
  geom_bar(position='stack')+
  xlab('Day in ICU')+
  ylab('Number of patients')+
  scale_fill_manual('Anticoagulants', values=c('indianred1','cyan3','grey'), labels=c('No','Yes','Missing'))+
  g.theme+
  theme(axis.text.x = element_text(size=5))+ # reduce label size
  facet_wrap(~facet, scales='free_y')
bplot
```

The y-axes on the two panels differ because there are far fewer patients in the ECMO group.

## Anticoagulants type in the last 24 hours

```{r, fig.width=8}
to_plot_anti = filter(to_plot, eotd_anticoagulants_type!='')
colours = gg_color_hue(10)
bplot = ggplot(data=to_plot_anti, aes(x=factor(day), fill=factor(eotd_anticoagulants_type)))+
  geom_bar(position='stack')+
  xlab('Day in ICU')+
  ylab('Number of patients')+
  scale_fill_manual(NULL, values=colours)+
  g.theme+
  theme(axis.text.x = element_text(size=5))+ # reduce label size
  facet_wrap(~facet, scales='free_y')
bplot
```

The y-axes on the two panels differ because there are far fewer patients in the ECMO group.

## Haemorrhagic complication

```{r, fig.width=8}
to_plot = mutate(to_plot, eotd_haemorrhagic_complication_plot = ifelse(is.na(eotd_haemorrhagic_complication), 2, eotd_haemorrhagic_complication)) # make extra category for missing
bplot = ggplot(data=to_plot, aes(x=factor(day), fill=factor(eotd_haemorrhagic_complication_plot)))+
  geom_bar(position='stack')+
  xlab('Day in ICU')+
  ylab('Number of patients')+
  scale_fill_manual('Haemorrhagic\ncomplication', values=c('indianred1','cyan3','grey'), labels=c('No','Yes','Missing'))+
  g.theme+
  theme(axis.text.x = element_text(size=5))+ # reduce label size
  facet_wrap(~facet, scales='free_y')
bplot
```

The y-axes on the two panels differ because there are far fewer patients in the ECMO group.

## Source of Haemorrhagic Complication 

```{r, results='asis'}
to_table = filter(to_plot, eotd_haemorrhagic_complication==1)
freq(to_table$eotd_haemorrhagic_complication_source, cumul=FALSE, report.nas = FALSE)
```

Just for the `r nrow(to_table)` days with a haemorrhagic complication in the first 21 days. The total number of patients is `r length(unique(to_table$pin))`.

### Number of patients with "central nervous system"

```{r, include=FALSE}
to_count = filter(to_plot, eotd_haemorrhagic_complication==1,
                  eotd_haemorrhagic_complication_source == 'Central nervous system'
                  ) %>%
  select(pin, complication_stroke) %>%
  unique()
n_patients_cns = nrow(to_count)
```

The previous table is the number of days and so contains repeated days from the same patients. The number of unique patients with at least one day of "central nervous system" is `r n_patients_cns`.

The table below shows whether these `r n_patients_cns` patients also had stroke (using the complication data).

```{r, results='asis'}
freq(to_count$complication_stroke, cumul=FALSE, report.nas = TRUE)
```

# Stroke

## Number of strokes

```{r, results='asis'}
tab_stroke = with(patients, freq(complication_stroke, round.digits=0, report.nas	=FALSE, cumul = FALSE))
tab_stroke
```

## Stroke types


```{r}
# add ggupset
cplot = ggplot(filter(patients, complication_stroke=='Yes'), aes(x = stroke_type)) +
    geom_bar(aes(y=..count..), fill='seagreen') +
    theme_bw() +
    xlab("") +
    ylab("Number of patients") +
    scale_x_upset( )+
    g.theme+
  theme(plot.margin = margin(t = 5, r = 0, b = 0, l = 110, unit = "pt"))
cplot
```

This is a new variable which was added in August 2020. Up to two options for the stroke type variable could be ticked. The plot is just for the `r sum(patients$complication_stroke=='Yes', na.rm=TRUE)` patients with a stroke complication.

The third most common response is missing stroke type, which reflects the fact that the data is being added retrospectively. The most common response with complete data is "Intraparenchymal haemorrhage". 

## Association between a cannula in the internal jugular vein and stroke/ICH/death

The table below shows the association between insertion site and stroke complication.

```{r}
tab = group_by(ecmo_patients, ecmo_drainage_cannula_insertion_site, complication_stroke) %>%
  tally() %>%
  group_by(ecmo_drainage_cannula_insertion_site) %>%
  mutate(percent = prop.table(n)*100,
    cell = paste(n, ' (', roundz(percent, digits=0), ')', sep='')) %>%
  ungroup() %>%
  mutate(ecmo_drainage_cannula_insertion_site = ifelse(ecmo_drainage_cannula_insertion_site=='', 'Missing', ecmo_drainage_cannula_insertion_site ),
    complication_stroke = ifelse(is.na(complication_stroke), 'Missing', as.character(complication_stroke))) %>%
  select(ecmo_drainage_cannula_insertion_site, complication_stroke, cell) %>%
  spread(complication_stroke, cell)
#
table_header <- data.frame(
  col_keys = c( 'ecmo_drainage_cannula_insertion_site', 'Missing', 'No', 'Yes' ),
  h1 = c('', rep("Stroke complication", 3)),
  h2 = c('Drainage cannula insertion site', 'Missing', 'No', 'Yes'),
  stringsAsFactors = FALSE )
#
ftab = flextable(tab) %>%
 set_header_df(mapping = table_header, key = "col_keys" ) %>% # add header
 merge_h(i=1, part = "header") %>% # merge "To" column headers
  theme_box() %>%
  autofit()
ftab
```

The table shows the number of patients and the row percentage.